\mutagen: Coverage-Guided, Property-Based Testing using Exhaustive Type-Preserving Mutations

### 2021-04-08 papers/oopsla-21/main.tex
Added references, general pass

%% PBT and generation of interesting data for free
%
Random Property-Based Testing (RPBT) is a popular technique for finding bugs
using executable properties encoding the expected behavior of our system.
%
Exemplified by frameworks Haskell's \quickcheck \cite{ClaessenH00} and the
abundant ports of this tool in other programming languages
\cite{hughes2003erlang, papadakis2011proper, bulwahn2012new,
  denes2014quickchick}, the most common approach is to instantiate the testing
properties using randomly generated inputs.

Users of \quickcheck (and RPBT in general) are well aware that one of the
biggest challenges while using these frameworks is to provide suitable random
data generators that are used to instantiate the testing properties.
%
In some extreme cases, writing highly-tuned random generators capable of
exercising every part of a complex system in a reasonable basis can take several
thouthands person-hours of trial and error \cite{lampropoulos2019coverage}.

To alleviate this, there exist several techniques to automatically synthesize
random data generators by reyfing the static information present in the
codebase, e.g., data type definitions, application public interfaces (APIs),
etc. \cite{GriecoCB16, DBLP:conf/haskell/MistaRH18, Mista2019GeneratingRS,
  DuregardJW12, Lampropoulos2017, Bendkowski2017}.
%
These techniques, however, are unable to synthesize random generators capable of
producing data satisfying complex invariants not easily derivable from the
codebase.
%
Randomly generating valid programs is common example of this problem, where
developers are forced to put a substantial effort writing complex generators by
hand \cite{Palka11, perenyi2020stack, yang2011finding}.
%
Moreover, automatically synthesized random generators offer particularly poor
results when the testing properties are constrained by sparse preconditions.
%
In this scenario, most of the generated tests are simply discarded before they
can be used to test the main components of our system.

@@

Instead of continuously generating random invalid test cases from scratch,
\fuzzchick keeps a queue of interesting previously executed test cases that can
be mutated using type-preserving transformations in order to produce new ones.
%
Intuitively, mutating an interesting test case in a small way (at the data
constructor level) has a much higher probability of producing a new interesting
test case than generating a new one from scratch using a na\"ive generator.
%
\fuzzchick is likely to preserve the semantic structure of the mutated data, as
mutations are applied directly at the data type level -- the random AST in case
of generating code.

This tool relies on execution traces to distinguish which test cases resulted
interesting and are therefore worth mutating --- a prolific technique known
\emph{coverage-guided fuzzing}, popularized by tools like \emph{AFL} \cite{afl}.
%
Mutated test cases are considered interesting for mutation only when they
produce new execution traces -- any other test case is simply thrown away.

@@

%% Problems of FuzzChick: mutations are superficial, long queues, power
%% scheduler
%
In this work, we take a close look at \fuzzchick for opportunities to improve.
%
Notably, we establish several aspects of its implementation that can be revised
--- some of these already being identified by the authors of \fuzzchick on their
original work.
%
In particular:
%
\begin{inparaenum}
\item \emph{the type-preserving mutations produced by this tool are superficial
  and rely heavily on randomness to produce diverse mutants},
\item \emph{using normal queues can cause large delays when interesting values
  are enqueued frequently}, and
\item \emph{using a power schedule to assing a certain mutation energy to each
  interesting test case requires fine tuning and can be hard to generalize to
  work well under different testing scenarios.}
\end{inparaenum}

Moreover, we replicated the IFC stack machine case study bundled with \fuzzchick
and observed a noticeable limitation:
%
\emph{after repeating each experiment 30 times, \fuzzchick was only able to find
  5 out of the 20 injected bugs with a 100\% efficacy, the hardest one being
  found only around 13\% of the time after an hour of testing.}
%
Despite this limitation, \citeauthor{lampropoulos2019coverage} show that
\fuzzchick is far better than using normal random testing coupled with na\"ive
random generators by comparing the mean-time-to-failure (MTTF) against
\emph{QuickChick}.
%
However, we believe that a proper evaluation ought to use the failure rate as an
important metric when comparing RPBT tools.

%% Our solution
In this work we introduce \mutagen, a RPBT tool implemented in Haskell that
build upon the basic idea behind \fuzzchick, i.e., code instrumentation and
type-preserving mutations.
%
\mutagen is introduced in detail in Section \ref{sec:mutagen}.

%% Exhaustive mutations
Unlike \fuzzchick, mutations in \mutagen are applied \emph{exhaustively}.
%
This is, given an interesting, previously executed test case, our tool
precomputes and schedules every possible type-presserving mutation that can be
applied to it.
%
This approach has two inherent advantages when compared to \fuzzchick.
%
On one hand, mutations do not rely on randomness to be generated.
%
Instead, every subexpression of the input test case is mutated on the same
basis, which guarantees that no interesting mutation is left behind due to
randomness.
%
On the other hand, scheduling mutations exaustively eliminates the need of a
heuristic power schedule to a given mutation energy to each input test
case.

Notably, \emph{scheduling mutations exhaustively within an input test case does
  not mean that mutations are exhaustive by themselves.}
%
Our tool distinguishes two kinds of mutations, those that can be computed
purely, yielding a single mutated value; and those that can be obtained
non-deterministically.
%
Pure mutations encode transformations that can be done by swapping data
constructor variants or rearranging subexpressions, whereas non-deterministic
(random) mutations are useful to represent transformations over large
enumeration types, e.g., numbers, characters, etc.
%
In this setting, random mutations let us selectively escape the exhaustiveness
of our tool by randomly sampling a small random generator a reduced number of
times.
%
This way, our tool avoids, for instance, mutating a given number into every
other number in the 32 or 64 bits range.

%% Heuristics to improve testing loop
Additionally, the testing loop of our tool incorporates two heuristics that help
finding bugs faster and more reliably (Section \ref{sec:heuristics}).
%
In first place, \mutagen uses first-in first-out (FIFO) scheduling with priority
to enqueue interesting test cases for mutation.
%
This scheduling algorithm is indexed by the novelty of each new test case with
respect to the ones already seen.
%
In this light, test cases that discover new parts of the code earlier during
execution are given a higher priority while picking the next mutated test case.
%
Additionally, our scheduling lets the testing loop jump back and forth between
scheduled test cases as soon as a new more interesting one is enqueued for mutation,
eliminating potential delays when the mutation queues grow large.

The second heuristic is used to tune the number of times our tool samples random
mutants, i.e., those corresponding to large enumeration types as described above.
%
For this, we keep track of how often we generate interesting test cases.
%
If this frequency suddenly stalls, \mutagen automatically resets the testing
loop increasing the amount of times we sample such mutants.
%
This automatically finds a suitable value for this external parametere on the
fly.

%% Validation
To validate our ideas, we use two main case studies (described in detail in
Section \ref{sec:casestudies}): the IFC stack machine used by
\citeauthor{lampropoulos2019coverage} on their original work, and a WebAssembly
engine written in Haskell.
%
In both case studies, we additionally compare the effect of the heuristics
implemented on top of the base testing loop of \mutagen.
%
In the IFC stack machine case study, we compare \fuzzchick against different
variants of our tool, where our results (Section \ref{sec:evaluation}) indicate
that \mutagen outperforms \fuzzchick both in terms of time to failure and
failure rate.
%
On the other hand, our WebAssembly case study shows the performance of our tool
in a more realistic scenario.
%
There, \mutagen is capable of reliaby finding 15 planted bugs in the validator
and interpreter, as well as 3 previously unknown bugs that flew under the radar
of the existing unit test suite of this engine.
%
All in all, our evaluation encompass more than 600 hours of cpu time and suggest
that testing mutants exhaustively can be an appealing technique for finding bugs
reliably without sacrificing speed.

@@

Property-based testing is a powerful technique for finding bugs without having
to write test cases by hand.
%
Originally introduced by \citet{quickcheck} in tandem with the first version of
\quickcheck, this technique focuses on aiming the developer's efforts into
testing systems via executable properties using randomly generated inputs.
%
Moreover, tools like \quickchick and Isabelle's \quickcheck demonstrate
that PBT can also be used in the formal verification realm.
%
There, one can quickly spot bugs in system specifications before directing the
efforts into pointlessly trying to prove bogus propositions.

@@

\noindent In this work we focus solely on the first two elements introduced
above, namely the testing properties and the random data generators used to feed
them.
%
Printers and shrinkers, for the most part, can be obtained automatically using
generic programming capabilities present in the compiler, and although being
crucial for the testing process as a whole, their role becomes irrelevant when
it comes to \emph{finding} bugs.

@@

\noindent The type parameter \texttt{a} indicates that trees can be instantiated
using any other type as payload, so the value \texttt{Leaf Bool} has type
\texttt{Tree Bool}, whereas the value \texttt{Branch (Leaf 1) 2 (Leaf 3)} has
type \texttt{Tree Int}.
%
Then, we can define tree reflections using a simple recursive function that
pattern matches against the two possible constructors, inverting the order of
the subtrees whenever it encounters a branch:

@@

With our simple specification in place, the last missing piece is a random
generator of trees.
%
In \quickcheck, this is usually done via the type class mechanism, instantiating
the \texttt{Arbitrary} type class, providing a random generator as the
implementation of the overloaded \texttt{arbitrary} operation:

@@

To alleviate the problem of testing properties with sparse preconditions while
using simple automatically derived random generators, \fuzzchick introduces
\emph{coverage-guided property-based testing} (CGPT) by enhancing the testing
process with two key characteristics:
%
\begin{inparaenum}
\item \emph{target code instrumentation}, to capture execution information from
  each test case; and
\item \emph{high-level, type-preserving mutations}, to produce syntactically
  valid test cases by altering existing ones at the datatype level.
\end{inparaenum}

@@

In addition, high-level mutators are better suited for producing syntactically
valid mutants, avoiding the time wasted by using generic low-level mutators that
act at the ``serialized'' level and know very little about the structure of the
generated data, thus producing syntactically broken mutants most of the time.
%
This grammar-aware mutation technique has shown to be quite useful when fuzzing
systems accepting structurally complex inputs.
%
Tools like Criterion, XSmith and LangFuzz use existing grammars to tailor the
generic mutators to the specific input structure used by the system under test.
%
In \fuzzchick, external grammars are not required.
%
The datatypes used by the inputs of the testing properties already describe the
structure of the random data we want to mutate in a concrete manner, and
specialized mutators acting at the data constructor level can be automatically
derived directly from their definition.

@@

Mutators in \fuzzchick are no more than specialized random generators,
parameterized by the original input to be mutated.
%
They use a simple set of mutation operations that are randomly applied at the
datatype level.
%
In simple terms, these operations encompass
%
\begin{inparaenum}
\item \emph{shrinking the value}, replacing its outtermost data constructor with
  one that contains a subset of its fields, reusing existing subexpressions;
\item \emph{growing the value}, replacing its outtermost data constructor with
  one that contains a superset of its fields, reusing existing subexpressions
  and generating random ones when needed;
\item \emph{returning a subexpression of the same type}; and
\item \emph{mutating recursively}, applying a mutation operation over an
  immediate subexpression.
\end{inparaenum}

@@

Fig. \ref{fig:fuzzchick:mutator} illustrates a \fuzzchick mutator for our
previously used \texttt{Tree} data type example.
%
Since trees are parametric, for clarity this definition is also parameterized by
a mutator for the payload (\texttt{mutate\_a}), although this can be abstracted
away using the type class system.

@@

\paragraph{Mutators distribution:}
%
if we inspect the mutator defined in Fig. \ref{fig:fuzzchick:mutator}, there are
two compromises that the authors of \fuzzchick adopted for the sake of
simplicity.
%
On one hand, deep recursive mutations are very unlikely, since their probability
decreases multiplicatively with each recursive call.
%
For instance, mutating a subexpression that lies on the third level of a
\texttt{Tree} happens with a probability smaller than $(1/6)^3 = \sim 0.0046$,
i.e., when applying a recursive mutation (with probability $1/6$) three
consecutive times.
%
This only worsens as the type of the mutated value becomes more complex.
%
Hence, \fuzzchick mutators can only be effectively used to transform to shallow
data structures, potentially excluding interesting applications that might
require producing deeper valid values, e.g., programs, network protocols
interactions, etc.
%
Ideally, mutations should be able to happen on every subexpression of the input
seed in a reasonable basis.

On the other hand, using random generators to produce neeeded subexpressions
when growing data constructors can be dangerous, as we are introducing the very
same ``uncontrolled'' randomness that we wanted to mitigate in the first place!
%
If the random generator produces an invalid subexpression (something quite
likely), this might just invalidate the whole mutated test case.
%
We believe that growing data constructors needs to be done carefully.
%
For instance, by using just a minimal piece of data to make the overall mutated
test case type correct.
%
If that mutated test case to be interesting, that subexpression can always be
mutated later.

\paragraph{Enqueuing mutation candidates:}
%
\fuzzchick uses two single queues for keeping valid and discarded mutation
candidates.
%
Whenever a new test case is found interesting, it is placed \emph{at the end of
  its corresponding queue}.
%
If this test case happens to have discovered a whole new portion of the target
code, it will not be further mutated until the rest of the queue ahead of it
gets processed.
%
This can limit the effectiveness of the testing loop if the queues tend to grow
more often than they tend to shrink, as interesting mutation candidates can get
buried at the end of a long queue that only exercises the same portion of the
target code.
%
In the extreme case, they might not processed at all within the testing budget.
%
Ideally, one would like a mechanism that prioritizes mutating test cases that
discovers new portions of the code right away, and that is capable of jumping
back and forth from mutation candidates whenever this happens.
%
We show in Section \ref{sec:heuristics} how this can be achieved by analyzing
the execution information in order to prioritize test cases with novel execution
traces.

@@

In this section we describe the base ideas behind \mutagen, our CGPT tool
written in Haskell.
%
Despite the choice of implementation language, we remark that the ideas
presented in this work extend beyond Haskell and functional programming
languages in general.
%
%% Later, Section \ref{sec:heuristics} introduces heuristics that help finding bugs
%% faster in certain testing scenarios.

In constrast with \fuzzchick, \mutagen does not employ a power schedule to
assign energy to mutable candidates.
%
In turn, it resorts to mutate them in an exhaustive and precise basis, where
%
\begin{inparaenum}
  \item each subexpression of a mutation candidate is associated with a set of
    deterministic mutations, and
  \item for every mutable subexpression, each one of these mutations is
    evaluated \emph{exactly once}.
\end{inparaenum}
%
%% There is a small exception to this rules that we will introduce soon.

This idea is inspired by exhaustive bounded testing tools like SmallCheck (in
Haskell) or Korat (in Java), that produce test cases exhaustively.
%
In simple words, such tools work by enumerating all possible values of the
datatypes used in the testing properties, and then executing them from smaller
to larger until a a bug is found, or a certain size bound is reached.
%
The main problem with this approach is that the space of all possible test cases
often grows exponentially as we increment the size bound, and the user
experiences what it looks like ``hitting a wall'', where no larger test cases
can be evaluated until we exhausted all the immediately smaller ones \cite{feat}.
%
To alleviate this problem, such tools often rely on heuristics to prune the
search space --- this is discussed in Section \ref{sec:related}.

@@

Not to be confused by these tools, in \mutagen we do not enumerate all possible
test cases exhaustively.
%
\mutagen uses random generators to find interesting initial seeds, and only then
proceeds to execute all possible mutations.
%
Moreover, like in \fuzzchick, the testing loop of our tool automatically filters
the test cases that are worth mutating by keeping only those that discover new
execution paths in the code under test.

@@

To describe \mutagen's testing loop, we first need to introduce the mechanism
used for testing mutations exhaustively.
%
In contrast to \fuzzchick, where mutators are parameterized random generators,
in \mutagen we define mutations as the set of mutants that can be obtained by
altering the input value at the top-level (the root).
%
In Haskell, we define mutations as:

@@

Pure mutants are used most of the time, and encode simple deterministic
transformations over the outtermost data constructor of the input --- recursive
mutations will be introduced soon.
%
These tranformation can:
%
\begin{inparaenum}
\item return an immediate subexpression of the same type as the input;
\item swap the outtermost data constructor with any other constructor of the
  same type, reusing existing subexpressions; and
\item rearrange and replace fields using existing ones of the same type.
\end{inparaenum}
%
To illustrate this, Fig. \ref{fig:mutagen:mutator} outlines a mutator for the
\texttt{Tree} data type.
%
Notice how this definition simply enumerates mutants that transform the
outtermost data constructor, hence no recursion is needed here.
%
Moreover, notice how a default value (\texttt{def}) used to fill the subtrees
when transforming a leaf into a branch.
%
This value corresponds to the smallest value we can construct in order for the
mutant to be type-correct.
%
In practice (\texttt{def = Leaf def}), where the inner \texttt{def} is the
smallest value of the payload --- we use the type class system to abstract this
complexity away in our implementation.
%
Using a default smallest value is again inspired by exhaustive bounded testing
tools, and avoids introducing unnecesary randomness when growing data
constructors.

@@

At this point, the reader might consider:
%
what about large enumeration types like integers or characters?
%
Does \mutagen transform them exaustively too?
%
\emph{Certainly not.}
%
Random mutants serve as a way to break exhaustiveness when mutating values of
large enumeration types.
%
Instead of trying every possible integer or character, we resolve in using a
random generator and sample a small amount of values from it --- the precise
amount is a tunable parameter of \mutagen.
%
This way, a mutator for integers simply becomes:

@@

So far we defined mutations that transform the root of the input.
%
Now it is time to apply these mutations to every subexpression as well.
%
To do so, we will use two functions that can also be derived from the data type
definition.

In first place, a function \texttt{positions} traverses the mutable candidate
and builds a rose tree of mutable positions.
%
These positions are essentially a list of indices encoding the path from the
root to every mutable subexpression.
%
For instance, the positions of a \texttt{Tree} are computed as follows:

@@

\noindent Later, given a desired position to mutate within a candidate, we
define another function \texttt{inside} that precisely finds the subexpression
corresponding to it and applies the mutation.
%
A slightly simplified version of this function for the \texttt{Tree} datatype is
as follows:

@@

In this section we introduce two heuristics implemented on top of the base
testing loop of our tool.
%
\mutagen enables them all by default, although they can be individually disabled
by the user if deemed appropriate.

@@

Obtaining good random generators automatically from static information (e.g.,
grammars or datatype definitions) has been tackled from different angles.

>>>>

%% PBT and generation of interesting data for free
%
Property-Based Testing (PBT) is a popular technique for finding bugs using
executable properties that encode the expected behavior of our systems.
%
Exemplified by frameworks like Haskell's \quickcheck \cite{ClaessenH00} and the
abundant ports of this tool in other programming languages
\cite{hughes2003erlang, papadakis2011proper, bulwahn2012new,
  denes2014quickchick}, the most common PBT approach is to instantiate the
testing properties using randomly generated inputs.

Users of \quickcheck (and PBT in general) are well aware that one of the biggest
challenges while using these frameworks is to provide suitable random data
generators needed to instantiate the testing properties.
%
In some extreme cases, writing highly-tuned random generators capable of
exercising every part of a complex system in a reasonable basis can take several
thouthand person-hours of trial and error \cite{lampropoulos2019coverage}.

To alleviate this, there exist several techniques to automatically synthesize
random data generators by reyfing the static information present in the
codebase, e.g., data type definitions, application public interfaces (APIs),
etc. \cite{GriecoCB16, DBLP:conf/haskell/MistaRH18, Mista2019GeneratingRS,
  DuregardJW12, Lampropoulos2017, Bendkowski2017}.
%
These techniques, however, are unable to synthesize random generators capable of
producing data satisfying complex invariants not easily derivable from the
codebase.
%
Randomly generating valid programs is common example of this problem, where
developers are forced to put a substantial efforts in writing specialized random
generators by hand \cite{Palka11, perenyi2020stack, yang2011finding}.
%
Moreover, automatically synthesized random generators offer particularly poor
results when the testing properties are constrained by sparse preconditions.
%
In this scenario, most of the generated tests are simply discarded before they
can be used to test the main components of our system.

@@

Instead of continuously generating random invalid test cases from scratch,
\fuzzchick keeps a queue of interesting previously executed test cases that can
be mutated using type-preserving transformations in order to produce new ones.
%
Intuitively, mutating an interesting test case in a small way (at the data
constructor level) has a much higher probability of producing a new interesting
test case than generating a new one from scratch using a na\"ive generator.
%
In this light, \fuzzchick is likely to preserve the semantic structure of the
mutated data, as mutations are applied directly at the data type level, e.g.,
the program's AST in case of generating code.

This tool relies on execution traces to distinguish which test cases resulted
interesting and are therefore worth mutating --- a prolific technique known as
\emph{coverage-guided fuzzing} and popularized by tools like \emph{AFL}
\cite{afl}.
%
Mutated test cases are considered interesting for mutation only when they
produce new execution traces --- any other test case is simply thrown away.

@@

%% Problems of FuzzChick: mutations are superficial, long queues, power
%% scheduler
%
In this work, we take a close look at \fuzzchick for opportunities to improve.
%
Notably, we establish several aspects of its implementation that can be revised
--- some of these already being identified by the authors of \fuzzchick in their
original work.
%
In particular:
%
\begin{inparaenum}
\item \emph{the type-preserving mutations produced by this tool are superficial
  and rely heavily on randomness to produce diverse mutants},
\item \emph{using normal queues can cause large delays when interesting values
  are enqueued frequently}, and
\item \emph{the mechanism used to assing a certain mutation energy to each
  interesting test case requires fine tuning and can be hard to generalize to
  work well under different testing scenarios.}
\end{inparaenum}

Moreover, we replicated the IFC stack machine case study bundled with \fuzzchick
and observed a noticeable limitation:
%
\emph{after repeating each experiment 30 times, \fuzzchick was only able to find
  5 out of the 20 injected bugs with a 100\% efficacy, the hardest one being
  found only around 13\% of the time after an hour of testing.}
%
Despite this limitation, \citeauthor{lampropoulos2019coverage} show that
\fuzzchick is far better than using normal random testing coupled with na\"ive
random generators by comparing the mean-time-to-failure (MTTF) against
\emph{QuickChick}.
%
However, we believe that a proper evaluation ought to use the failure rate as an
important metric when comparing PBT tools.

%% Our solution
In this work we introduce \mutagen, a PBT tool implemented in Haskell that
builds upon the basic ideas behind \fuzzchick, i.e., code instrumentation and
type-preserving mutations.
%
\mutagen is introduced in detail in Section \ref{sec:mutagen}.

%% Exhaustive mutations
Unlike \fuzzchick, mutations in \mutagen are applied \emph{exhaustively}.
%
This is, given an interesting, previously executed test case, our tool
precomputes and schedules every type-presserving mutation that can be applied to
it.
%
This approach has two inherent advantages when compared to \fuzzchick.
%
On one hand, mutations do not rely on randomness to be generated.
%
Instead, every subexpression of the input test case is mutated on the same
basis, which guarantees that no interesting mutation is left behind due to
randomness.
%
On the other hand, scheduling mutations exaustively eliminates the need of a
heuristic that assigns a given mutation energy to each input test case.

Notably, \emph{scheduling mutations exhaustively within an input test case does
  not mean that mutations are exhaustive by themselves.}
%
Our tool distinguishes two kinds of mutations, those that can be computed
purely, yielding a single deterministic mutated value; and those that can be
obtained non-deterministically.
%
On one hand, pure mutations encode transformations that swap data constructors
around, as well as return or rearrange subexpressions.
%
Non-deterministic (random) mutations, on the other hand, are useful to represent
transformations over large enumeration types, e.g., numbers, characters, etc.
%
This mechanism let us selectively escape the exhaustiveness of our tool by
randomly sampling a small random generator a reduced number of times.
%
This way, our tool avoids, for instance, mutating a given number into every
other number in the 32 or 64 bits range.

%% Heuristics to improve testing loop
Additionally, the testing loop of our tool incorporates two heuristics that help
finding bugs faster and more reliably (Section \ref{sec:heuristics}).
%
In first place, \mutagen uses first-in first-out (FIFO) scheduling with priority
to enqueue interesting test cases for mutation.
%
This scheduling algorithm is indexed by the novelty of each new test case with
respect to the ones already seen.
%
In this light, interesting test cases that discover new parts of the code
earlier during execution are given a higher priority.
%
Additionally, our scheduling lets the testing loop jump back and forth between
mutable test cases as soon as a new more interesting one is enqueued for
mutation, eliminating potential delays when the mutation queues grow large.

The second heuristic is used to tune the number of times our tool samples random
mutants, i.e., those corresponding to large enumeration types as described above.
%
For this, we keep track of how often we generate interesting test cases.
%
If this frequency suddenly stalls, \mutagen automatically resets the testing
loop increasing the amount of times we sample such mutants.
%
This automatically finds a suitable value for this external parameter on the
fly.

%% Validation
To validate our ideas, we use two main case studies (described in detail in
Section \ref{sec:casestudies}): the IFC stack machine used by
\citeauthor{lampropoulos2019coverage} on their original work, and a WebAssembly
engine written in Haskell.
%
In both case studies, we additionally compare the effect of the heuristics
implemented on top of the base testing loop of \mutagen.
%
In the IFC stack machine case study, we compare \fuzzchick against different
variants of our tool, where our results (Section \ref{sec:evaluation}) indicate
that \mutagen outperforms \fuzzchick both in terms of time to failure and
failure rate.
%
On the other hand, our WebAssembly case study shows the performance of our tool
in a more realistic scenario.
%
There, \mutagen is capable of reliably finding 15 planted bugs in the validator
and interpreter, as well as 3 previously unknown bugs that flew under the radar
of the existing unit test suite of this engine.
%
All in all, our evaluation encompass more than 600 hours of cpu time and suggest
that testing mutants exhaustively can be an appealing technique for finding bugs
reliably without sacrificing speed.

@@

Property-based testing is a powerful technique for finding bugs without having
to write test cases by hand.
%
Popularized by Haskell's \quickcheck, this technique focuses on aiming the
developer's efforts into testing systems via executable properties using
randomly generated inputs.
%
%% Moreover, tools like \quickchick and Isabelle's \quickcheck demonstrate
%% that PBT can also be used in the formal verification realm.
%% %
%% There, one can quickly spot bugs in system specifications before directing the
%% efforts into pointlessly trying to prove bogus propositions.

@@

\noindent In this work we focus solely on the first two elements introduced
above, namely the testing properties and the random data generators used to feed
them.
%
Printers and shrinkers, for the most part, can often be obtained automatically
using generic programming capabilities present in the compiler, and although
being crucial for the testing process as a whole, their role becomes irrelevant
when it comes to \emph{finding} bugs.

@@

\noindent The type parameter \texttt{a} indicates that trees can be instantiated
using any other type as payload, so the value \texttt{Leaf True} has type
\texttt{Tree Bool}, whereas the value \texttt{Branch (Leaf 1) 2 (Leaf 3)} has
type \texttt{Tree Int}.
%
Then, we can define tree reflections using a simple recursive function that
pattern matches against the two possible constructors, inverting the order of
the subtrees whenever it encounters a branch:

@@

The last missing piece is a random generator of trees.
%
In \quickcheck, this is commonly done via the type class mechanism
\cite{jones1997type}, instantiating the \texttt{Arbitrary} type class with a
random generator as the implementation of the overloaded \texttt{arbitrary}
operation:

@@

To alleviate the problem of testing properties with sparse preconditions while
using automatically derived random generators, \fuzzchick introduces
\emph{coverage-guided, property-based testing} (CGPT), a technique that enhances
the testing process with two key characteristics:
%
\begin{inparaenum}
\item \emph{target code instrumentation}, to capture execution information from
  each test case; and
\item \emph{high-level, type-preserving mutations}, to produce syntactically
  valid test cases by altering existing ones at the data type level.
\end{inparaenum}

@@

In addition, high-level mutators are better suited for producing syntactically
valid mutants, avoiding the time wasted by using generic low-level mutators that
act at the ``serialized'' level and know very little about the structure of the
generated data, thus producing syntactically broken mutants most of the time.
%
This grammar-aware mutation technique has shown to be quite useful when fuzzing
systems accepting structurally complex inputs.
%
Tools like LangFuzz \cite{holler2012fuzzing}, Superion \cite{wang2019superion},
XSmith \cite{xsmith} use existing grammars to tailor the mutators to the
specific input structure used by the system under test.
%
In \fuzzchick, external grammars are not required.
%
The data types used by the inputs of the testing properties already describe the
structure of the random data we want to mutate in a concrete manner, and
specialized mutators acting at the data constructor level can be automatically
derived directly from their definition.

@@

Mutators in \fuzzchick are no more than specialized random generators,
parameterized by the original input to be mutated.
%
They use a simple set of mutation operations that are randomly applied at the
data type level.
%
In simple terms, these operations encompass
%
\begin{inparaenum}
\item \emph{shrinking the value}, replacing its outtermost data constructor with
  one that contains a subset of its fields, reusing existing subexpressions;
\item \emph{growing the value}, replacing its outtermost data constructor with
  one that contains a superset of its fields, reusing existing subexpressions
  and generating random ones when needed;
\item \emph{returning a subexpression of the same type}; and
\item \emph{mutating recursively}, applying a mutation operation over an
  immediate subexpression.
\end{inparaenum}

@@

Fig. \ref{fig:fuzzchick:mutator} illustrates a \fuzzchick-like mutator for our
previously used \texttt{Tree} data type example.
%
Since trees are parametric, for clarity this definition is also parameterized by
a mutator for the payload (\texttt{mutate\_a}), although this can be abstracted
away using the type class system.

@@

\paragraph{Mutators distribution:}
%
if we inspect the mutator defined in Fig. \ref{fig:fuzzchick:mutator}, there are
two compromises that the authors of \fuzzchick adopted for the sake of
simplicity.
%
On one hand, deep recursive mutations are very unlikely, since their probability
decreases multiplicatively with each recursive call.
%
For instance, mutating a subexpression that lies on the fourth level of a
\texttt{Tree} happens with a probability smaller than $(1/6)^3 = \sim 0.0046$,
i.e., when applying a recursive mutation (with probability $1/6$) three
consecutive times.
%
This only worsens as the type of the mutated value becomes more complex.
%
Hence, \fuzzchick mutators can only be effectively used to transform to shallow
data structures, potentially excluding interesting applications that might
require producing deeper valid values, e.g., programs, network protocols
interactions, etc.
%
Ideally, mutations should be able to happen on every subexpression of the input
seed in a reasonable basis.

On the other hand, using random generators to produce neeeded subexpressions
when growing data constructors can be dangerous, as we are introducing the very
same ``uncontrolled'' randomness that we wanted to mitigate in the first place!
%
If the random generator produces an invalid subexpression (something quite
likely), this might just invalidate the whole mutated test case.
%
We believe that growing data constructors needs to be done carefully.
%
For instance, by using just a minimal piece of data to make the overall mutated
test case type correct.
%
If that mutated test case turns to be interesting, that subexpression can always
be mutated later.

\paragraph{Enqueuing mutation candidates:}
%
\fuzzchick uses two single queues for keeping valid and discarded mutation
candidates.
%
Whenever a new test case is found interesting, it is placed \emph{at the end of
  its corresponding queue}.
%
If this test case happens to have discovered a whole new portion of the target
code, it will not be further mutated until the rest of the queue ahead of it
gets processed.
%
This can limit the effectiveness of the testing loop if the queues tend to grow
more often than they tend to shrink, as interesting mutation candidates can get
buried at the end of a long queue that only exercises the same portion of the
target code.
%
In the extreme case, they might not processed at all within the testing budget.
%
Ideally, one would like a mechanism that prioritizes mutating test cases that
discovers new portions of the code right away, and that is capable of jumping
back and forth from mutation candidates whenever this happens.
%
%% We show in Section \ref{sec:heuristics} how this can be achieved by analyzing
%% the execution information in order to prioritize test cases with novel execution
%% traces.

@@

In this section we describe the base ideas behind \mutagen, our CGPT tool
written in Haskell.
%
Despite the choice of implementation language, the ideas presented in this work
extend beyond Haskell and functional programming languages in general.
%
%% Later, Section \ref{sec:heuristics} introduces heuristics that help finding bugs
%% faster in certain testing scenarios.

In constrast with \fuzzchick, \mutagen does not employ a power schedule to
assign energy to mutable candidates.
%
In turn, it resorts to mutate them in an exhaustive and precise basis, where
%
\begin{inparaenum}
  \item each subexpression of a mutation candidate is associated with a set of
    type-preserving mutations, and
  \item for every mutable subexpression, each one of these mutations is
    evaluated \emph{exactly once}.
\end{inparaenum}
%
%% There is a small exception to this rules that we will introduce soon.

This idea is inspired by exhaustive bounded testing tools like SmallCheck
\cite{runciman2008smallcheck} or Korat \cite{boyapati2002korat}, that produce
test cases exhaustively.
%
In simple words, such tools work by enumerating all possible values of the
data types used in the testing properties, and then executing them from smaller
to larger until a a bug is found, or a certain size bound is reached.
%
The main problem with this approach is that the space of all possible test cases
often grows exponentially as we increment the size bound, and the user
experiences what it looks like ``hitting a wall'', where no larger test cases
can be evaluated until we exhausted all the immediately smaller ones
\cite{DuregardJW12}.
%
To alleviate this problem, such tools often rely on heuristics to prune the
search space based on detecting unevaluated subexpressions --- this is discussed
in Section \ref{sec:related}.

@@

Not to be confused by these tools, in \mutagen we do not enumerate all possible
test cases exhaustively.
%
Our tool uses random generators to find interesting initial seeds, and when it
finds them, only then proceeds to schedule all possible mutations.
%
Moreover, like in \fuzzchick, the testing loop of our tool automatically filters
the test cases that are worth mutating by keeping only those that discover new
execution paths in the code under test.

@@

To describe \mutagen's testing loop, we first need to introduce the mechanism
used for testing mutations exhaustively.
%
In contrast to \fuzzchick, where mutators are parameterized random generators,
in our tool we define mutations as the set of mutants that can be obtained by
altering the input value at the top-level (the root).
%
In Haskell, we define mutations as:

@@

Pure mutants are used most of the time, and encode simple deterministic
transformations over the outtermost data constructor of the input --- recursive
mutations will be introduced soon.
%
These tranformations can:
%
\begin{inparaenum}
\item return an immediate subexpression of the same type as the input;
\item swap the outtermost data constructor with any other constructor of the
  same type, reusing existing subexpressions; and
\item rearrange and replace fields using existing ones of the same type.
\end{inparaenum}
%
To illustrate this, Fig. \ref{fig:mutagen:mutator} outlines a mutator for the
\texttt{Tree} data type.
%
Notice how this definition simply enumerates mutants that transform the
outtermost data constructor, hence no recursion is needed here.
%
Moreover, notice how a default value (\texttt{def}) used to fill the subtrees
when transforming a leaf into a branch.
%
This value corresponds to the simplest value we can construct in order for the
mutant to be type-correct.
%
In practice (\texttt{def = Leaf def}), where the inner \texttt{def} is the
simplest value of the payload --- we use the type class system to abstract this
complexity away in our implementation.
%
Using a small value is again inspired by exhaustive bounded testing tools, and
avoids introducing unnecesary randomness when growing data constructors.

@@

%% At this point, the reader might consider:
%% %
%% what about large enumeration types like integers or characters?
%% %
%% Does \mutagen transform them exaustively too?
%% %
%% \emph{Certainly not.}
%
Random mutants, on the other hand, serve as a way to break exhaustiveness when
mutating values of large enumeration types --- or any other type the user might
want to introduce random mutations to.
%
Instead of mutating into every other possible number or character, we resolve in
using a random generator and sample a small amount of values from it --- the
precise amount is a tunable parameter of \mutagen.
%
This way, a mutator for integers simply becomes:

@@

So far we have defined mutations that transform only the root of the input.
%
Now it is time to apply these mutations to every subexpression as well.
%
To do so, we will use two functions that can be derived from the data type
definition.

In first place, a function \texttt{positions} traverses the input and builds a
rose tree of mutable positions.
%
These positions are essentially a list of indices encoding the path from the
root to every mutable subexpression.
%
For instance, the positions of a \texttt{Tree} are computed as follows:

@@

\noindent Later, given a desired position to mutate within a test case, we
define another function \texttt{inside} that finds the subexpression
corresponding to it and applies the mutation.
%
A slightly simplified version of this function for the \texttt{Tree} data type is
as follows:

@@

In this section we introduce two heuristics implemented on top of the base
testing loop of our tool.
%
\mutagen enables them all by default, although they can be individually disabled
by the user if deems appropriate.

@@

Obtaining good random generators automatically from static information (e.g.,
grammars or data type definitions) has been tackled from different angles.
